---
title: "frescalo"
author: "Colin Harrower, Jonathan Yearsley and Oliver Pescott"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The frescalo package contains an R implementation of Frescalo (FREquency SCAling LOcal) 
modeling methodology invented by Dr Mark O. Hill [(Hill, 2011)](https://doi.org/10.1111/j.2041-210X.2011.00146.x) to aid with analysis of biological record data. 
The methodology attempts to correct for spatial and temporal variation in recorder effort that is 
common in biological recording data (i.e. species occurrence data).


## Installation

You can install the development version of frescalo from [GitHub](https://github.com/colinharrower/frescalo) with:
```{r install_pkg, eval = FALSE}
# install.packages("pak")
pak::pak("colinharrower/frescalo")
```


## Introduction

Frescalo attempts to correct for spatial and temporal variation in recorder effort by analysing biological records data for a suite of species, often within a related  taxonomic group, and making inference of the recording effort based on the relative recording frequencies of these species. Frescalo uses the recording frequency of sets of **benchmark species**, species believed to be relatively ubiquitous and stable in occurrence, to determine recording effort. Defining such **benchmark species** across a wide area would be difficult so instead frescalo uses sets of **benchmark species** that are unique to localised **neighbourhoods**. 
In frescalo a **neighourhood** represent a set of locations that are biologically or ecologically similar and or located in relative proximity to each other. In the simplest case neighbourhoods can be defined purely on proximity, however incorporating biologically or ecological similarity into the definition and weighting of neighbourhoods can typically improve the results. 

The recording effort for each of these neighbourhoods is determined by calculating the weighted local species frequencies for each species in that neighbourhood and then from these species frequencies calculating a neighborhood frequency. The local species frequencies for that neighbourhood are then systematically adjusted by a neighbourhood sampling-effort modifier to inflate or deflate the species frequency curves until the neightbourhood frequency achieves a specified standard value (Phi).

## Data required

### Species occurrence data
Frescalo uses biological recording data where the original data has been summarised to unique combinations of time period, location and species. The choice of the spatial resolution for the locations and the time period chosen are up to the user but ideally these should chosen to allow an adequate level of recording at sites within the time periods.

Frescalo was designed with biological atlas data in mind and as such has often been applied at typical atlas scales, i.e. 10km^2^ square **sites** and multi-year atlas recording time periods (.e.g. 1980-1995, 2000-2015). Frescalo has however been sucessfully applied at finer spatial and or temporal resolutions, i.e. 1km^2^ or 2km^2^ sites and annual time periods.

### Neighbourhood weights

In addition to the species occurrence data Frescalo requires a dataset specifiying the sites that comprise each individual neighbourhood, and their relative weightings, for each site. Typically neighbourhoods are defined by first selecting the `k` closest sites and then from these selecting the `n` sites showing the greatest similarities to the focal site. The package includes a function, `calc_neight_wts()` to create a neighbourhood weights dataset from a `sf spatial polygons` or `spatial points`object containing attribute fields upon with the similarities are to be calculated.


## Example

The package contains simulated data which can be used as an example and or to test the package. The simulated data comrpises two data objects `s` containing occurrence data and `d` containing neighborhood weightings. The occurrence data in `s` has already been summarized to the format required by frescalo, specifically a `data.frame` containing all unique combinations of time period (`time`), sites or locations (`location`) and species ids or names (`species`). The neighbourhood weights data specifies for each focal location (`location1`) which other sites form its "neighbourhood" (`location2`) and the relevant weighting score between the focal site and each respective neighbourhood site (`w`).

```{r view_input}
# Load the package
library(frescalo)

# View the first 10 rows of the test dataset s (occurrence data)
head(s,10)

# View the first 10 rows of the weights datasets (d)
head(d,10)
```

In this example dataset the species are identified using simplistic species IDs and the locations are identified using Ordnance Survey of Great Britain 10 km grid references, however the site and species identifiers can be any appropriate names or coded identifiers. The time periods can be identified as simple coded or numerical identifiers, as in the example data `d`, or using appropriate numerical values (e.g. mid-point of the time periods).


The main wrapper function `frescalo()` is used to fit apply the frescalo model to the data, returning a list comprised of 4`data.frames` `locs` with the location metrics, `freq` with original and re-scaled species frequencies, `trend` with tFactors for each species, `site_time` with the site by time recording effort metrics used for the adjustments. The outputs are equivalent to the `samples.txt`, `frequencies.out` and `trend.out` output files produced by Mark Hill's original fortran frescalo program.

```{r fit_fresc}
# Use test dataset (s) and weights data (d) included with the package
out_fres = frescalo(s,d,in_parallel = FALSE,filter_wts = TRUE)

# Show the names of the elements in out_fres
names(out_fres)

```

The location metrics from frescalo are found in the first `locs` element of the output list object and is a data.frame structured as follows:

1. **location** - The location ID/code
2. **nSpecies** - The number of species recorded
3. **phi_in** - The original neighbourhood frequency, or frequency-weighted mean frequecy, for that locations neighbourhood
4. **alpha** - The sampling effort modifier required to standardise the neighourhood frequency to Phi
5. **phi_out** - The value of neightbourhood frequecy after rescaling (this should match the value of Phi specified as an target)
6. **spnum_in** - The sum of neighbourhood frequencies before rescaling
7. **spnum_out** - The sum of neighbourhood frequencies after scaling, i.e. the estimated species richness
8. **iter** - The number of iterations for the algorithm to determine the alpha value required for the neightbourhood frequency to reach phi


```{r res_locs}
# Show location outputs from the frescalo results object
head(out_fres[["locs"]])

```


The species frequencies metrics from frescalo are found in the `freq`` element of the output object and is a data.frame structured as follows:

1. **location** - The location ID/code
2. **species** - The species ID/code/Name
3. **pres** - Whether the species was recording in that location (0 = not recorded, 1 = recorded)
4. **freq** - The frequency of species in the neighbourhood
5. **freq_1** - The frequency of the species in the neighbourhood after rescaling, i..e the estimated probability of occurrence
6. **rank** - The rank of that species frequency in the neighbourhood
7. **rank_1** - The rescaled rank, defined as the rank/estimated species richness
8. **benchmark** - Specifying whether the species was considered a benchmark species for the neighbourhood (1 = benchmark species, 0 = non-benchmark species)


```{r res_freq}
# Show the rescaled species frequecy outputs from the frescalo results object
head(out_fres[["freq"]])

```

The temporal tFactor metrics from frescalo are found in the `trend`` element of the output object and is a data.frame structured as follows:

1. **species** - The species ID/code/Name
2. **time** - The ID/code/value specifying the time period
3. **tFactor** - The frescalo time factor, i.e. the estimated relatively frequency of species at the time
4. **StDev** - The standard deviation of the time factor


```{r res_trend}
# Show tFactor trend outputs from the frescalo results object
head(out_fres[["trend"]])

```

The site by time metrics from frescalo are found in the `site_time` element of the output object and is a data.frame structured as follows:

1. **location** - The location ID/code
2. **time** - The ID/code/value specifying the time period
3. **s_it** - 
4. **w** - 


```{r res_site_time}
# Show the site by time recording effort outputs from the frescalo results object
head(out_fres[["site_time"]])

```

```{r plot_tfactor}

# Plot tFactor for a specific species
  plot_tfactor(out_fres[["trend"]][which(out_fres[["trend"]]$species == "Species 1"),])

```


```{r plot_alpha}
library(sf)
# Merge location results from frescalo with spatial polygons for locations in example neighbourhoods (stored in d_locs)
  sf_locs = merge(d_locs, out_fres[["locs"]], by = "location")
# Plot alpha
  fres_map(sf_locs,zcol="alpha")
# Plot estimated species richness
  fres_map(sf_locs,zcol="spnum_out")


```


